The demographic costs of sexually antagonistic selection in partially selfing populations

### 2022-01-05 doc/ResponseLetter-R1.tex

%===========================================
% Document

\section*{}
\noindent To the Ed., AE, \& Editorial Board
\bigskip

\noindent Please consider this revision of our manuscript (MS \#60734), titled "The demographic costs of sexually antagonistic selection in partially selfing populations" for publication in \textit{The American Naturalist}. 
\bigskip

\noindent First of all, we want to thank the Ed., AE, and EiC for their thoughtful and forthright responses regarding the editorial delay on our manuscript. We completely understand that pandemic-related personal issues have been and continue to be extremely disruptive in these difficult times, and that editorial duties for our journals is a volunteer service. To our AE, Dr.~Orive, we sincerely hope that you and your family are well, and we truly appreciate the time and effort you took to evaluate our paper when you clearly had more imporant things to deal with.
\bigskip

@@

%%%%%%%%%%%%%%%%%%%%
\section*{Decision letter with our responses}

\subsection*{Responses to Editor Erol Ak\c{c}ay}

I must first apologize for the time it took to return this decision, and thank you for your patience. Even in the best of times, unexpected events and circumstances cause delays for reviewers and editors, and many of us are still operating under the continuing strain of the pandemic. In this case, I must also apologize to have added to the delay personally, which is because of a backlog that built up due to disruptions that happened in the last couple of months that ordinarily might not have led to huge issues, but after a year and a half of the pandemic seem to snowball out of control.

\begin{quote}
	{\itshape Thank you, again, for being so candid and forthright. We truly appreciate it, and sympathize with the growing challenges that have crept up on us all during the pandemic.}
\end{quote}

@@

As you will see in Dr. Orive’s excellent summary of the reviews, one issue that we feel you need to address is to make the paper more suitable for the general audience of the American Naturalist. To my mind, this has two parts: First, I think you can reduce the amount of space you devote to the general model in the main text (this can go into the SI as a generalization and also as a reference for the later three-stage model for Mimulus), as most of the biological conclusions actually come from the two-stage model which requires less notation and the construction of which is more easily explained. Although I personally appreciated the construction of the general model, I think for our non-theoretical readership, the cost/benefit ratio of reading through the general model (which is not used for the numerical calculations) is too great.

\begin{quote}
	{\itshape We agree that shifting the focus from deriving the general model to the specific two stage example will make the paper more accessible and shorten the Methods. We have now moved the development of the general model to the Online Supplementary Material (Supplement 1), and restructured the Methods in the main paper accordingly. }
\end{quote}

Second, like Reviewer 1 and Dr. Orive, I do think you should better connect your model to the broader topics of evolution in structured populations and sexual antagonism. That would give your paper a broader impact in terms of readership.

\begin{quote}
	{\itshape We agree entirely and have restructured our introduction accordingly. Please see our detailed responses to Dr.~Orive and R1 on this count.}
\end{quote}

@@

\begin{quote}
	{\itshape This is a very good and totally valid point. Honestly, we believe a separate paper is necessary to deal with the many consequences of density dependence for our model predictions. However, we also agree that sidestepping this issue completely isn't good enough. We have now made two main changes to address density dependence. First, have added a new {\bf Appendix B} where we briefly show how density dependence can be incorporated into the model, and reanalyze our {\itshape M. guttatus} example with a simple density-dependent version of the model. Second, we have expanded our methods, results, and especially the discussion to more fully address the implications of density-dependence, with reference to our modified figures and the new {\bf Appendix B} (see {\bf L.128--134; 444--447; 529--549}).}
\end{quote}

@@

Related to that, and without doing more modeling, one quantity of interest you can present is the actual growth rate, and not just the region of viability/inviability. In a world where we acknowledge density dependence as a possibility, how far above or below one growth rate is can be informative about whether you actually expect extinction or depressed densities.

\begin{quote}
	{\itshape This is great suggestion which inspired us to make several major changes. We now include a new illustrative figure ({\bf fig.~1}), and revised versions of figs. 1 \& 4 (now {\bf figs.~S1 \& 5}) which show heatmaps of absolute values for predicted $\lambda$ instead of just extinction boundaries. Unfortunately, when presenting $\lambda$ in this way, it is no longer possible to show multiple values of maximum fertility ($f$) on a single panel. In particular, our old fig. 1 now requires 18 panels to illustrate $\lambda$ for the same set of parameter combinations. We felt that this resulted in an unnecessarily large and complicated figure. For simplicity, we have decided to retain our original version of fig.1 (now {\bf fig. 2}) in the main text and include the modified version showing 18 heatmaps for $\lambda$ as a supplementary figure (now {\bf fig.~S1}). We hope that, with the new illustrative {\bf fig.~1}, the extinction boundaries presented in {\bf fig.~2} will be easier for readers to understand. However, if the editors and reviewers prefer the full heatmaps, we will happily swap {\bf fig.~2} with {\bf fig.~S1}. Lastly, we agree that the {\itshape M. guttatus} example is greatly improved by showing $\lambda$, and we now present the modified version of fig. 4 (now {\bf fig. 5}) in the main text}.
\end{quote}

@@

\noindent xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

\section*{Responses to Associate Editor Maria Orive}

Both reviewers found the work interesting and potentially to be of strong interest to the readership of AmNat, and I agree. However, Reviewer 1 brings up an important point regarding the framing of the work and placing it within a broader context of the literature on evolutionarily explicit structured models, which I think is important for the authors to do in order to make the paper accessible to and of interest for the overall readership of AmNat. There are other points that the reviewers bring up that are important to consider, but I think this one needs to be addressed carefully.

\begin{quote}
	{\itshape We can see now that in our efforts to appeal to both demographers and evolutionary biologists studying sexual conflict, we insufficiently discussed the relevant literature on both counts. We have restructured our introduction to better engage with these fields without lengthening it (see, in particular, {\bf L.43--52, L.58--67, L.76--85}). See also our specific responses to Reviewer 1. }
\end{quote}

@@

Reviewer 2 has some very nice suggestions regarding ideas for simplifying and clarifying the model description.

\begin{quote}
	{\itshape Agreed. Please see our sepecific responses to Reviewer 2 below.}
\end{quote}

In addition to the clarifications/edits and other points raised by the reviewers, I include a handful of minor edits/typos here of my own:
\bigskip

\noindent (1) Abstract, perhaps add a very general and short explanation of the term sexually antagonistic polymorphism.
\begin{quote}
	{\itshape We have added a brief general explanation of sexually antagonistic selection, which we felt was lacking in the first draft, and also allows us to more clearly link SA polymorphism with its' demographic consequences ({\bf L.2--6}).}
\end{quote}

@@

\noindent (2) Line 165; "fertility"
\begin{quote}
	{\itshape Fixed.}
\end{quote}

\noindent (3) Table 1, definition of inbreeding depression terms; it would be clearer to define each of the 4 terms separately, "Inbreeding depression terms for ovule viability, juvenile survival, adult survival, and transition rates respectively"
\begin{quote}
	{\itshape Fixed}
\end{quote}

\noindent (4) Figure 1; I generally prefer axes labels to have clear meanings, rather than just symbols – here, the vertical and horizontal axes give the female (vertical) vs. male (horizontal) selection on reproductive function.
\begin{quote}
	{\itshape Good point. We have changed our axes labels to be more descriptive }
\end{quote}

@@

\begin{quote}
	{\itshape Thank you again for your efforts and consideration in spite of these challenges. We really appreciate it.}
\end{quote}

\noindent xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

\subsection*{Reviewer \#1:}

The manuscript reports the derivation, analysis, and example application of an extended frequency-dependent matrix model to explore links between sexual antagonistic selection, population persistence, and the maintenance of heritable variation. The paper is very well written and is straightforward to follow. The derivations are correct, although I did not try to derive all the equations in the appendices. My one comment is how might environmental stochasticity in the rate of selfing and outcrossing impact general conclusions? 
\begin{quote}
	{\itshape We now briefly discuss this possibility in the Discussion ({\bf L.564--570}).}
\end{quote}

@@

The manuscript adds to a growing and important literature where eco-evolutionary feedbacks are examined in structured models, some of which incorporate evolution explicitly. It also contributes nicely to a body of empirical and theoretical work on sexually antagonistic selection.

\begin{quote}
	{\itshape Thank you for these encouraging words.}
\end{quote}

As written the manuscript will not appeal to the broad readership of the American Naturalist because the weakness of the manuscript is the author makes little attempt to fit the work into the broader relevant literature. Instead, the paper has been written for a small audience of experts interested in the demographic consequences of sexually antagonistic selection. As written, the work is suited to a journal such as Journal of Theoretical Biology. This is disappointing, and the authors do themselves no favors by writing for such a niche audience. The manuscript could be well-suited to the American Naturalist if the authors were to engage with the broader literature. This can be done by extending the introduction to link this work to the wider literature on evolutionarily explicit structured models, and to the large empirical literature on sexually antagonistic selection. If they are able to do this appropriately, I believe this paper would make a strong contribution to the American Naturalist.

\begin{quote}
	{\itshape This is a very important comment and we thank you for pointing out this major shortcoming. Our goal when crafting the first draft was to find a middle ground that would appeal to both demographers and people studying sexual conflict. It is clear that we didn't quite achieve that goal. Rather, by making compromises in the introduction we ended up lacking scholarship on both sides. To address your main concern, we have restructured the Introduction to more directly connect with the broader literature on evolutionarily explicit structured models (see {\bf L.58--67}) and now also provide a stronger link with the empirical sexually antagonistic selection literature (see {\bf L.43--51}). We hope these changes, along with a more streamlined presentation of SA selection (see {\bf L.34--39; 52--54}) succeed in broadening the appeal of our manuscript to the wider AmNat readership.}
\end{quote}

@@

\noindent xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

\subsection*{Reviewer \#2:}

I have read, with great interest, this manuscript analyzing the effects of sexually antagonistic selection in hermaphrodites at both the population genetics and population dynamics levels. As a representative (I guess) of the general readership of AmNat and also as a population dynamics specialist, I appreciated the description of the model, its parametrization, the way it is constructed and analyzed. Not being a population genetics specialist, it is harder for me to comment on that side of the interpretations, apart from generic comments and in the light of the model’s characteristics. Overall, my take is that this paper is interesting, well written, the model well built and the ideas quite novel; and that it therefore corresponds to what readers except to find in AmNat both in terms of content and quality.

\begin{quote}
	{\itshape Thank you for the very encouraging assessment.}
\end{quote}

@@

\noindent I have one main comment:

\noindent-- Discussion: Conclusions are drawn before limitations of the model are mentioned. However, it seems to me that it would be very important to discuss the effect of those limitations on the results themselves and therefore discuss the generality of the results. For instance, what is the potential effect of female dominance in the marriage function on the results and in particular on conclusion of lines 477-480.

\begin{quote}
	{\itshape While we feel that there is strong precedence for discussing the key results of a study before diving into the various caveats for with the analysis, yoru point is well taken that we need to more clearly acknowledge these limitations in the methods and direct readers to where these limitations are discussed. In our Methods section, we are now careful to point readers to the various Appendixes, Supplements, and places in the Discussion where these limitations are dealt with (e.g., see {\bf L.94--98, 103--109, 128--134, 183--186, 187--189, 196--197, 207--212, 274--275}), and we have added more explicit discussion of several of our key assumptions in the discussion ({\bf e.g., L.519--581}). We discuss both the causes and potential effects of incorporating male-limitation of reproductive success on {\bf L.550--561}}.
\end{quote}

@@

\noindent Apart from that, I have only minor comments, that I list below (roughly in decreasing order of importance):
\bigskip

\noindent -- Model: lines 114-195
\begin{itemize}
	\item The matrices H are mentioned before they are introduced. (Therefore, the parenthesis in lines 123-124 cannot be understood by the reader). And then are only defined in the following section	
	\begin{itemize}
		\item It would make sense to reorder the model presentation by presenting the building blocks (arguably U, F and H matrices, but I'll come back to H in the next point), what they do, what they are for; then provide some properties (F independent from selfing/crossing, H from stage, etc ...) and then go into the details of their construction
	\end{itemize}
	\item The main issue I have here, is that H is introduced, but then disappears from the output (see equations 12 and 13 for instance that use H but it does not appear, or the general output eq 9)
	\begin{itemize}
		\item This suggests combining F and H directly into an object.
	\end{itemize}
\end{itemize}

\begin{itemize}
	\item More generally, the matrix formulations of the model may be a bit hard to read for the general reader of AmNat. Getting rid of H (or to the contrary, using it all the way, thus simplifying eq 12 and 13 for instance) and not giving a name to matrix W (which is just a bunch of 0s 1s and 1/2s) are steps that can make it easier for the reader to follow.
\begin{quote}
	{\itshape {\bf \ul{Response to the above 5 points}}: As per the reviewers (and the editors) suggestion, we have moved most of the derivations of the general model into the supplementary materials, and now no longer use H or W in the main text. In doing so, we hope to shift the focus from mathematical derivations towards explaining the end result instead. However, we have made the requested changes to clarify the derivation in {\bf Online Supplement 1}. }
\end{quote}

@@

\end{itemize}

\noindent -- Line 19: I find it weird to write that the "process of adaptation can be impeded by ...changing environmental conditions".

\begin{itemize}
	\item One organism can be adapted to an environment (as in a series of environmental parameters) but also to changes in that environment (plasticity, genetic variance etc ...). It is actually from this consideration that a lot of studies on adaptation and adaptive dynamics.
	\item I think this sentence should be rewritten to make it clear that it is not changes in environmental conditions per se that impede the adaptive process but some of their characteristics (speed, range etc.)
\end{itemize}

\begin{quote}
	{\itshape These are good points, that was awkward wording on our part. While revising the introduction, we felt that much of this first paragraph was tangential to the key concepts we wanted to introduce. We have now re-written and shortened it to more quickly and concisely introduce several concepts which we feel are more relevant to the rest of the introduction ({\bf L.20--24}). The offending sentence has been removed.}
\end{quote}

@@

\noindent -- Line 23: The link between genetic constraints and genetic trade-offs should be clarified. Or rather a definition of each term provided, for the reader to understand this sentence. I suspect that, for many a reader of AmNat, genetic constraints and genetic trade-offs are equivalent concepts.

\begin{quote}
	{\itshape Agreed. We have rephrased to avoid the term "trade-off" which tends to over-simplify the concept in our opinion ({\bf L.23--24}).}
\end{quote}

@@

\noindent -- Line 30: "fitness trade-offs ... provide ...mechanism for the maintenance of genetic variation". This requires a few words of explanation. Is that a general statement about (fitness) trade-offs? or is it actually particular to SA trade-offs or to trade-offs between fixed heterogeneity components?
\begin{itemize}
	\item For instance, a fitness trade off between fertility at 3 years and fertility at 6 years will have no positive effect on genetic variance. To the contrary, it may lead the organism to evolve an optimal strategy accounting for overall fitness and the trade-offs, vis-à-vis investments in reproduction at those different stages in the life cycle. And as such this may lead to the selection of individuals with such a strategy and therefore to low genetic variance.
\end{itemize}

\begin{quote}
	{\itshape We have rephrased the first sentence of this paragraph to be more specific ({\bf L.25--26}), which we hope also addresses this comment.}
\end{quote}

@@

\noindent -- Line 264: this is an important point that is a bit unclear here, according to me. Could you clarify that what you call demographic rates are the transitions in the overall model and so include both demography and genetics? Clarify also that the model is not density dependent because it only depends on frequencies ... and maybe relate it to a general equation where A does not depend on n per se but on the q frequencies.

\begin{quote}
	{\itshape We have added a few lines to discuss frequency vs density dependence when we first introduce the population projection matrix in equation 3 ({\bf L.131--132}). In addition, we have tried to clarify to improve this section by discussing that the survival and fertility rates ("demographic rates") are density-indepedent, and that the allele frequency dynamics make the model frequency dependent ({\bf L.128--134}).}
\end{quote}

@@

- Line 35: SA selection arises also from non-genetic trade-offs, which makes the formulation of that sentence a bit weird.

\begin{quote}
	{\itshape We have rephrased this sentence to clarify and be more consistent with our language earlier in the preceding paragraphs ({\bf L.39--42}).}
\end{quote}

@@

- Lines 289-292: inbreeding depression is expressed as a function of C. However, C is not written in the equation but L which depends on C. C is the introduced parameter and so should appear. If the expression is too complicated then it should be described literally, but there is no need to introduce L here, in any case.

\begin{quote}
	{\itshape We now present the full expression involving $C$ as Eq(16).}
\end{quote}

@@

- Line 42: Define analogous here

\begin{quote}
	{\itshape This sentence has been rephrased.}
\end{quote}

- Line 54: maybe a few words to remind the reader about what dioecious and hermaphrodite organisms are.

\begin{quote}
	{\itshape We have added a brief parenthesis to remind readers what dioecy is ({\bf L.50}). Hermaphroditism is defined earlier in the paragraph, so an additional reference seemed redundant.}
\end{quote}

@@

- Line 80: just a few words to clarify "obligate outcrossing" (something like (no selfing allowed) maybe)

\begin{quote}
	{\itshape Fixed ({\bf L.85-86}).}
\end{quote}

- Line 233: f and f’ are called "adult fertilities". However, they correspond to that rate only for certain specific genotypes. Maybe clarify that by calling them "adult maximum fertilities" ?

\begin{quote}
	{\itshape Good point, fixed ({\bf L.169, 175, Table 2}).}
\end{quote}

@@

- Results: it took me a while to figure out Fig 1 (mainly because I was looking at it in B and W). I am just wondering whether it would be easier to read with only one f per chart. Maybe as a separate chart that illustrates the general shape of the parameter space for polym * extinction for only one C, one h and one f... then followed by these 6 graphs as they are now.

### 2022-01-05 doc/SA-Hermaphrodites-wDemography-R1.tex
Finalized revisions, response letter, etc.

\textcolor{blue}{When selection differs between the sexes, genes expressed by both males and females can experience sexually antagonistic (SA) selection, where beneficial alleles for one sex are deleterious for the other.} Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of such SA genetic polymorphisms, but these models have rarely considered the demographic consequences \textcolor{blue}{ of coexisting alleles with deleterious fitness effects in each sex.} In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying when SA polymorphisms are maintained by balancing selection {\itshape and} the population growth rate remains positive. Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations can sustain large regions of viable SA polymorphic space.

@@

%To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{different fitness components are negatively correlated} \citep{CharlesworthHughes2000, ConnallonHall2018}. 
\textcolor{blue}{A population's ability to adapt to its environment and persist in the long term depends on the genetic variation that it carries \citep{Fisher1930, Lewontin1975}. Yet, populations are heterogenous collections of individuals existing in a complex world where adaptation can be impeded by conflicting selection and gene-flow between different classes of individuals, or more generally, when different fitness components are negatively correlated \citep{Prout2000, CharlesworthHughes2000, ConnallonHall2018}.}

Genetic constraints on adaptation due to conflicting selection have particularly interesting evolutionary consequences \citep{ConnallonDebarreLi2018}. On one hand, they provide an effective mechanism for the maintenance of genetic variation, which shapes a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy, harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

@@

Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. Likewise, alleles with opposing fitness effects through male and female sex functions can cause SA selection in hermaphrodite populations, where individuals' fitness is determined jointly by their maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Moreover, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and polygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}. Overall, both theory and current empirical data suggest that there is ample scope for the maintenance of SA \textcolor{blue}{genetic variation} in both dioecious \textcolor{blue}{(separate-sexed)} and hermaphrodite populations \citep{BondurianskyChenoweth2009, Abbott2011, RuzickaESEB2020, WangBarrett2020}.}

@@

\textcolor{blue}{While significant efforts have been made to predict when SA polymorphisms should be maintained and to detect SA genetic variation from empirical data, less attention has been paid to the demographic consequences of SA selection \citep{MatthewsConnallon2019}}. Most population genetic theory, including SA selection theory, models the relative fitness of genotypes in populations of constant (and often infinite) size. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{which link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). For over 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. Theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic consequences of allele frequency changes due to selection. Various extensions of these theories have been made to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and the evolution of quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects}. This growing body of eco-evolutionary theory makes plain that demographic processes must be taken into account explicitly to fully understand how genetic variation for fitness is shaped, and maintained in natural populations \citep{metcalf2007evolutionary,coulson2006putting,rueffler2006evolutionary}.}

@@

The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few studies since have explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with SA selection and population dynamics, but did not analyze how demography influenced the maintenance of SA polymorphism. In this paper, we extend the model of \citet{deVriesCaswell2019a,deVriesCaswell2019b} to \textcolor{blue}{include both dioecious and hermaphroditic populations and jointly analyze the genetic (fixation of male- or female-beneficial alleles, or balanced polymorphism) and demographic (population persistence or extinction) outcomes of SA selection}. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata}).

@@

Simultaneous hermaphrodites can transmit genes to the next generation via both sperm/pollen and eggs/ovules, and have the potential to reproduce by a combination of self- and outcross-fertilizations. Maternal outcrossing involves receiving male gametes from other individuals in the population, while paternal outcrossing involves exporting male gametes to other individuals and fertilizing their ovules. Self-fertilization occurs when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female reproductive functions, we denote matrices or vectors relating to the male sex function with a prime. \textcolor{blue}{Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen, for reasons of analytic tractability. We present a more general model of genotype-specific self-fertilization rates in the Online Supplement, and briefly discuss implications of this approach in the Discussion.}

>>>>

\textcolor{blue}{When selection differs between the sexes, genes expressed by both males and females can experience sexually antagonistic (SA) selection, where beneficial alleles for one sex are deleterious for the other.} Classic population genetics theory has been fundamental to understanding \textcolor{blue}{how and when SA genetic variation can be maintained by balancing selection}, but these models have rarely considered the demographic consequences \textcolor{blue}{of coexisting alleles with deleterious fitness effects in each sex.} In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying when SA polymorphisms are maintained by balancing selection {\itshape and} the population growth rate remains positive. Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations can sustain large regions of viable SA polymorphic space.

@@

%To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{different fitness components are negatively correlated} \citep{CharlesworthHughes2000, ConnallonHall2018}. 
\textcolor{blue}{A population's ability to adapt to its environment and persist in the long term depends upon the genetic variation that it carries \citep{Fisher1930, Lewontin1975}. Yet, populations are heterogenous collections of individuals existing in a complex world where adaptation can be impeded by conflicting selection and gene-flow between different classes of individuals, or more generally, when different fitness components are negatively correlated \citep{Prout2000, CharlesworthHughes2000, ConnallonHall2018}.}

Genetic constraints on adaptation \textcolor{blue}{due to conflicting selection }have particularly interesting evolutionary consequences. On one hand, they provide an effective mechanism for the maintenance of genetic variation, which shapes a population's capacity for future adaptation \citep{Fisher1930, Prout2000, CharlesworthHughes2000, ConnallonHall2018}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy, harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

@@

Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. Likewise, alleles with opposing fitness effects on male and female sex functions can experience SA selection in hermaphrodite populations, where individual fitness is determined jointly by maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, sex-differences in directional selection are commonly detected in phenotypic selection studies \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, suggesting that many genetic variants contributing to quantitative traits have SA effects. Moreover, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and polygenic signals of SA selection are evident in several recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}. Overall, both theory and current empirical data suggest that there is ample scope for the maintenance of SA \textcolor{blue}{genetic variation} in both dioecious \textcolor{blue}{(separate-sexed)} and hermaphrodite populations \citep{BondurianskyChenoweth2009, Abbott2011, RuzickaESEB2020, WangBarrett2020}.}

@@

\textcolor{blue}{While significant efforts have been made to predict when SA polymorphisms can be maintained and to detect SA genetic variation from empirical data, the demographic consequences of SA selection have often been overlooked \citep{MatthewsConnallon2019}}. Most population genetic theory, including SA selection theory, models the relative fitness of genotypes in populations of constant (and often infinite) size. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{that link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). For over 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. Theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic consequences of allele frequency changes due to selection. Various extensions of these theories have been made to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and quantitative trait evolution \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects}. This growing body of eco-evolutionary theory makes plain that demographic processes must be taken into account to fully understand how genetic variation for fitness is shaped in natural populations \citep{metcalf2007evolutionary,coulson2006putting,rueffler2006evolutionary}.}

@@

The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few studies since have explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with SA selection and population dynamics, but did not analyze how demography influenced SA polymorphism. In this paper, we extend the model of \citet{deVriesCaswell2019a,deVriesCaswell2019b} to \textcolor{blue}{include both dioecious and hermaphroditic populations and jointly analyze the genetic (fixation/polymorphism) and demographic (persistence/extinction) outcomes of SA selection}. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata}).

@@

Simultaneous hermaphrodites can transmit genes to the next generation via both sperm/pollen and eggs/ovules, and can potentially reproduce by a combination of self- and outcross-fertilization. Maternal outcrossing involves receiving male gametes from other individuals in the population, while paternal outcrossing involves exporting male gametes to fertilize other individuals' ovules. Self-fertilization occurs when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female reproductive functions, we denote matrices or vectors relating to the male sex function with a prime. \textcolor{blue}{Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen. We present a more general model of genotype-specific self-fertilization rates in the Online Supplement, and briefly discuss implications of this approach in the Discussion.}

Individuals are jointly classified by life-cycle stage ($1, \ldots, \omega$), genotype ($1, \ldots, g$), and whether they were produced by self- or outcross fertilization (denoted by $S$ and $X$ superscripts, respectively; see Table \ref{tab:Terms} for a full description of terms included in the model). We include how individuals were produced in the state description because individuals produced by selfing might experience reduced survival, growth, or maturation rates due to inbreeding depression. \textcolor{blue}{This approach makes the common simplifying assumption that all individuals in the population experience the same severity of inbreeding depression (e.g., \citealt{Charlesworth1987, Charlesworth2009, Charlesworth2010, JordanConnallon2014}). We address potential consequences of relaxing this assumption, and how our model can be extended to do so, in the discussion. }

%In reality, the severity of inbreeding depression will not only be a function of whether an individual is produced by selfing but also of how many consecutive generations of inbreeding have occurred in their lineage (e.g., \citealt{kelly1999response, kelly2007mutation}). However, tracking lineage-specific inbreeding histories is beyond the scope of this paper, and we therefore make the common assumption
